Primer on Production Functions and Economic Growth

### 2014-11-29 Articles/Prod Func Primer/ProdFuncPrimer.tex
Updates to primer.

Continuing the cooking analogy, consider producing toast ($Y$) 
with a fatty substance~(butter or margarine).
%
\begin{equation} \label{eq:toast}
  Y = a B + b M
\end{equation}
%
The linear model (Equation~\ref{eq:toast}) indicates that butter~($B$) and margarine~($M$) 
can be used interchangably as the fatty substance.
Butter~($B$) and margarine~($M$) are perfect substitutes.%
  \footnote{
  Butter and margarine may be perfect substitutes for all but the most discerning toast consumers.
  }%

@@

$\rho_1$ indicates the elasticity 
of substitution~($\sigma_1$) between capital~($k$) and labor~($l$).
The elasticity of substitution 
between capital~($k$) and labor~($l$) is related to $\rho_1$ and given by 
%
\begin{equation} \label{eq:sigma_1}
 \sigma_1 = \frac{1}{1+\rho_1} \; .
\end{equation}
%
Solving for $\rho_1$ gives
%
\begin{equation} \label{eq:rho_1}
  \rho_1 = \frac{1 - \sigma_1}{\sigma_1} \; .
\end{equation}

@@

Figure~\ref{fig:CES_isoquants} shows the simplifications from Table~\ref{tab:ces_simplifications}
graphically.
%
\begin{figure}[!ht]
\centering\
\includegraphics[width=0.8\linewidth]{figure/CESIsoquants.png}
\caption[CES isoquants]{CES isoquants (lines of constant output). 
$K$ is capital and $N$ is labor.~\citep{Klump:2011aa}}
\label{fig:CES_isoquants}
\end{figure}

@@

An additional production function, the Linear Exponential (LinEx%
    \footnote{
    The name ``LinEx'' comes from the form of the equation:
    production is \emph{linear} in energy~($e$) and \emph{exponential}
    in ratios of the factors of production~($k$, $l$, and $e$).
    }%
) function is given by K\"{u}mmel in several places as
%
\begin{equation} \label{eq:LinEx-kummel}
  y = e \exp{\left[ a \left( 2 - \frac{l + e}{k} \right)  + a c \left( \frac{l}{e} - 1 \right)\right]}
\end{equation}
%
An alternative presentation that affords easier interpretation is
%
\begin{equation} \label{eq:LinEx}
  y = A e \; ; \; A \equiv \mathrm{e}^{a\left[2 \left(1 - \frac{1}{\rho_k} \right) 
    + c \left(\vphantom{\frac{1}{\rho_k}}\rho_l - 1 \right)\right]} \; .
\end{equation}
%
The LinEx function was derived from thermodynamic considerations
\citep{Kummel:1980wx,Kummel:1982vy,Kummel:1985vz,Kummel:1989tf,Kummel:2002tx,Kummel:2010vz}. 
A key feature of the derivation involves
constraints among capital~($k$), labor~($l$), and energy~($e$).
These constraints were discussed in Section~\ref{sec:Leontief} above.%
  \footnote{
  As an aside, I'm not sure the constraints 
  described in Section~\ref{sec:Leontief} are the right ones.
  }%

@@

The LinEx model does, however, include terms that represent
efficiencies among capital, labor, and energy, in the form of 
the ratios $\rho_k$ and $\rho_l$ which are defined by
%
\begin{equation} \label{eq:rho_k}
  \rho_k \equiv \frac{k}{(\nicefrac{1}{2})(l+e)}
\end{equation}
%
and
%
\begin{equation} \label{eq:rho_l}
  \rho_l \equiv \frac{l}{e}
\end{equation}
%
and represent 
%
(\ref{eq:rho_k}) capital stock increase relative to the average of labor and energy (capital deepening) and 
(\ref{eq:rho_l}) labor increase relative to energy.
The effect of energy ($e$) on output ($y$) is enhanced if either
%
\begin{itemize}
\item{capital stock ($k$) has increased more than the average 
      of labor ($l$) and energy ($e$), i.e. if $\rho_k >$ 1, or}
\item{labor ($l$) has increased more than energy ($e$), i.e. if $\rho_l >$ 1.}
\end{itemize}
%
An economy that increases 
capital stock ($k$) and labor ($l$) at the same rate as energy ($e$) will have
capital and labor efficiency ratios ($\rho_k$ and $\rho_l$, respectively) of unity.
In that case, economic output ($y$) increases at the same rate as energy consumption ($e$).
According to the LiNex model, an economy that increases 
capital stock ($k$) without a commensurate increase in the average of labor
and energy [$(\nicefrac{1}{2})(l+e)$] will experience an increase in output ($y$) in excess 
of its increase of energy consumption ($e$), because $\rho_k > 1$ and
%
\begin{equation}
    \mathrm{e}^{a \left[2 \left(1 - \frac{1}{\rho_k} \right) 
    + c \left(\vphantom{\frac{1}{\rho_k}}\rho_l - 1 \right)\right]}
	> 1 \; .
\end{equation}
%
Similarly, an economy benefits by increasing labor ($l$) 
without a commensurate increase in energy ($e$), because $\rho_l > 0$. 
Thus, $\rho_k$ is an 
efficiency of using additional labor~($l$) and energy~($e$) to make 
additional capital stock~($k$) available to the economy, 
and $\rho_l$ is an efficiency of using additional energy~($e$) to make 
additional labor~($l$) available to the economy. 
Increases in $\rho_k$ and $\rho_l$ provide upward pressure on economic output~($y$),
as the only factor of production~($e$) is used more efficiently.

The parameter $a$ represents the overall importance of efficiencies $\rho_k$ and $\rho_l$, 
and the parameter $c$ represents the relative importance 
of the labor efficiency term~($\rho_l$) compared to the capital stock efficiency term~($\rho_k$).

@@

For the Leontief, linear, Cobb-Douglas, and CES production functions, 
output elasticities are constant with respect to time.
However, for the LinEx function, 
the values of coefficients $a$ and $c$ imply time-dependent
values of $\alpha$, $\beta$, and $\gamma$ as given by the following equations~\citep{Warr:2012cg}:

\begin{equation} \label{eq:LINEX_alpha}
  \alpha \equiv \frac{k}{y}\frac{\partial y}{\partial k} = a \left( \frac{l + e}{k} \right) \;,
\end{equation}

\begin{equation} \label{eq:LINEX_beta}
  \beta \equiv \frac{l}{y} \frac{\partial y}{\partial l} = a \left[ c \left( \frac{l}{e}\right) - \frac{l}{k} \right] \;,
\end{equation}

\noindent{}and

\begin{equation} \label{eq:LINEX_beta_2}
  \gamma \equiv \frac{e}{y} \frac{\partial y}{\partial e} 
         = 1 - a \left[ \frac{e}{k} + c \frac{l}{e} \right] 
         = 1 - \alpha - \beta \;.
\end{equation}

>>>>

Continuing the cooking analogy, consider producing toast ($Y$) 
with a fatty substance~(butter, $B$, or margarine, $M$).
%
\begin{equation} \label{eq:toast}
  Y = a B + b M
\end{equation}
%
The linear model (Equation~\ref{eq:toast}) indicates that butter~($B$) and margarine~($M$) 
can be used interchangably as the fatty substance.
Butter~($B$) and margarine~($M$) are perfect substitutes.%
  \footnote{
  Butter and margarine are perfect substitutes for all but the most discerning toast consumers.
  }%

@@

$\rho_1$ indicates the elasticity 
of substitution~($\sigma_1$) between capital~($k$) and labor~($l$).
The elasticity of substitution~($\sigma_1$)
between capital~($k$) and labor~($l$) is related to $\rho_1$ and given by 
%
\begin{equation} \label{eq:sigma_1}
 \sigma_1 = \frac{1}{1+\rho_1} \; .
\end{equation}
%
Solving for $\rho_1$ gives
%
\begin{equation} \label{eq:rho_1}
  \rho_1 = \frac{1 - \sigma_1}{\sigma_1} \; .
\end{equation}

@@

Figure~\ref{fig:CES_isoquants} shows the simplifications from Table~\ref{tab:ces_simplifications}
graphically.
%
\begin{figure}[!ht]
\centering\
\includegraphics[width=0.8\linewidth]{figure/CESIsoquants.png}
\caption[CES isoquants]{CES isoquants (lines of constant output, $Y$). 
$K$ is capital and $N$ is labor. From \citet{Klump:2011aa}.}
\label{fig:CES_isoquants}
\end{figure}

@@

An additional production function, the Linear Exponential (LinEx%
    \footnote{
    The name ``LinEx'' comes from the form of the equation:
    production is \emph{linear} in energy~($e$) and \emph{exponential}
    in ratios of the factors of production~($k$, $l$, and $e$).
    }%
) function is given by K\"{u}mmel in several places~\citep{Kummel:1980wx,Kummel:1982vy,
Kummel:1985vz,Kummel:1989tf,Kummel:2002tx,Kummel:2010vz} as
%
\begin{equation} \label{eq:LinEx-kummel}
  y = e \exp{\left[ a \left( 2 - \frac{l + e}{k} \right)  + a c \left( \frac{l}{e} - 1 \right)\right]} \; .
\end{equation}
%
An alternative presentation that affords easier interpretation is
%
\begin{equation} \label{eq:LinEx}
  y = A e \; ; \; A \equiv \mathrm{e}^{a\left[2 \left(1 - \frac{1}{\rho_k} \right) 
    + c \left(\vphantom{\frac{1}{\rho_k}}\rho_l - 1 \right)\right]} \; .
\end{equation}
%
The LinEx function was derived from thermodynamic considerations. 
A key feature of the derivation involves
constraints among capital~($k$), labor~($l$), and energy~($e$).
These constraints were discussed in Section~\ref{sec:Leontief} above.%
  \footnote{
  As an aside, I question whether the constraints 
  described in Section~\ref{sec:Leontief} are the right ones.
  }%

@@

The LinEx model does, however, include terms that represent
efficiencies among capital, labor, and energy, in the form of 
the ratios $\rho_k$ and $\rho_l$ which are defined by
%
\begin{equation} \label{eq:rho_k}
  \rho_k \equiv \frac{k}{(\nicefrac{1}{2})(l+e)}
\end{equation}
%
and
%
\begin{equation} \label{eq:rho_l}
  \rho_l \equiv \frac{l}{e}
\end{equation}
%
and represent 
%
(\ref{eq:rho_k}) capital stock increase relative to the average of labor and energy (capital deepening) and 
(\ref{eq:rho_l}) labor increase relative to energy.
The effect of energy ($e$) on output ($y$) is enhanced when $A > 1$,
which occurs if either
%
\begin{itemize}
\item{capital stock ($k$) has increased more than the average 
      of labor ($l$) and energy ($e$), i.e. if $\rho_k >$ 1 or}
\item{labor ($l$) has increased more than energy ($e$), i.e. if $\rho_l >$ 1.}
\end{itemize}
%
An economy that increases 
capital stock ($k$) and labor ($l$) at the same rate as energy ($e$) will have
capital and labor efficiency ratios ($\rho_k$ and $\rho_l$, respectively) of unity.
In that case, economic output ($y$) increases at the same rate as energy consumption ($e$).
According to the LiNex model, an economy that increases 
capital stock ($k$) without a commensurate increase in the average of labor
and energy [$(\nicefrac{1}{2})(l+e)$] will experience an increase in output ($y$) in excess 
of its increase of energy consumption ($e$), because $\rho_k > 1$ and
%
\begin{equation}
    \mathrm{e}^{a \left[2 \left(1 - \frac{1}{\rho_k} \right) 
    + c \left(\vphantom{\frac{1}{\rho_k}}\rho_l - 1 \right)\right]}
	> 1 \; .
\end{equation}
%
Similarly, an economy benefits by increasing labor ($l$) 
without a commensurate increase in energy ($e$), because $\rho_l > 1$. 
Thus, $\rho_k$ is an 
efficiency of using additional labor~($l$) and energy~($e$) to make 
additional capital stock~($k$) available to the economy, 
and $\rho_l$ is an efficiency of using additional energy~($e$) to make 
additional labor~($l$) available to the economy. 
Increases in $\rho_k$ and $\rho_l$ provide upward pressure on economic output~($y$),
as the only factor of production~($e$) is used more efficiently.

The parameter $a$ represents the overall importance of efficiencies $\rho_k$ and $\rho_l$, 
and the parameter $c$ represents the relative importance 
of labor efficiency~($\rho_l$) compared to capital stock efficiency~($\rho_k$).

@@

For the Leontief, linear, Cobb-Douglas, and CES production functions, 
output elasticities are constant with respect to time.
However, for the LinEx function, 
the values of coefficients $a$ and $c$ imply time-dependent
values of $\alpha$, $\beta$, and $\gamma$ as given by the following equations~\citep{Warr:2012cg}:
%
\begin{equation} \label{eq:LINEX_alpha}
  \alpha \equiv \frac{k}{y}\frac{\partial y}{\partial k} = a \left( \frac{l + e}{k} \right) \;,
\end{equation}

\begin{equation} \label{eq:LINEX_beta}
  \beta \equiv \frac{l}{y} \frac{\partial y}{\partial l} = a \left[ c \left( \frac{l}{e}\right) - \frac{l}{k} \right] \;,
\end{equation}
%
\noindent{}and
%
\begin{equation} \label{eq:LINEX_beta_2}
  \gamma \equiv \frac{e}{y} \frac{\partial y}{\partial e} 
         = 1 - a \left[ \frac{e}{k} + c \frac{l}{e} \right] 
         = 1 - \alpha - \beta \;.
\end{equation}
%
The LinEx model assumes constant returns to scale.
